---
title: "Project1"
author: "Sylvia Baeyens"
date: "9/20/2021"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r packages, echo= FALSE, include= FALSE}
#including necessary libraries
library(data.table)
library(dplyr)
library(tidyverse)
library(leaflet)
library(R.utils)
library(ggplot2)
library(stringr)
```

# 1a. Reading in Data Files
```{r data-read, cache= TRUE}
Data2004 = data.table::fread("ad_viz_plotval_data2004.csv")
Data2019 = data.table::fread("ad_viz_plotval_data2019.csv")
```

# 1b. Summarizing Data
```{r}
# for 2005 Data
dim(Data2004)
head(Data2004)
tail(Data2004)

# for 2020 Data
dim(Data2019)
head(Data2019)
tail(Data2019)
```

The 2004 data has 19233 rows & 20 columns. The 2019 data has 53086 rows & 20 columns. There is clearly much more data from 2019. Some variables of interest include the Date (character variable), the Site ID (integer variable), the daily Mean PM2.5 Concentration (numeric variable), the state and county (both character variables), and the site latitude and longitude (both numeric variables.) 

# 2. Combine the two data sets & variable name clean-up
```{r}
#combining data sets
#got stuck on adding the Year variable after combination, so did it manually before
Data2004$Year = 2004
Data2019$Year = 2019
DataTotal = rbind(Data2004, Data2019)


DataTotal = DataTotal %>%
  rename(
     PM2_5 = "Daily Mean PM2.5 Concentration",
     lat = "SITE_LATITUDE",
     lon = "SITE_LONGITUDE",
     city = "Site Name"
  )
```

# 3. Creating Map in Leaflet
```{r}
#assigning color palette
temp.pal = colorNumeric(c('navyblue','lightblue'), domain = DataTotal$Year)

leaflet(DataTotal) %>%
  addProviderTiles('CartoDB.Positron') %>%
  addCircles(
    lat = ~lat, lng=~lon, 
    label = ~paste0(round(Year,2), 'Year'), color = ~ temp.pal(Year),
    opacity=1, fillOpacity=1
    ) 

unique(Data2004$`Site ID`)
unique(Data2019$`Site ID`)
```
Most sites are along the coast of California and clustered around big cities. There are a few stations located in Central and Northern CA. It is clear that there were more stations in 2019 than in 2004 which is verified by checking to see the unique amount of stations in both years.

# 4a. Check for Missing or Improbable Data
```{r}

DataTotal[,table(is.na(PM2_5))]
summary(DataTotal$PM2_5)
conc = DataTotal[order(PM2_5)]
head(conc)
tail(conc)

```
There are NO missing values for PM2.5. However, there are negative numbers, as shown by a min of -2.2, which are implausible. Most of the negative PM 2.5 concentrations were measured at the Tahoe City-Fairway Drive Station in 2019. PM 2.5 concentrations are considered unhealthy above 35.4 ug/m3. However, some quick research showed that a concentration of 251 (the maximum) is not unheard of, but very rare and dangerous. Most of the extremely high PM2.5 measurements occurred in Yosemite NP in 2004. 

# 4b. Correct for Improbably Data
```{r}
dim(DataTotal)
DataTotal <- DataTotal[PM2_5>=0]
summary(DataTotal$PM2_5)
dim(DataTotal)
```
There were 283 instances of negative PM2.5 concentration values, which have now been removed.


# 5. Explore the Main Question of Interest at State, County & City Level

Main question of interest: Have daily concentrations of PM2.5 decreased over the last 15 years?

# 5a. At the state level!
```{r}
ggplot(DataTotal, mapping=aes(x=1, y= PM2_5)) +
    geom_boxplot(aes(color= Year)) +
    facet_grid(~Year) +
    theme(legend.position = "none")
```
The mean PM2.5 Concentration in 2004 has higher than that of 2019, which means that concentration levels have in fact decreased when considering the entire state of California. 2019 does have a smaller range of concentrations, while 2004 had a few extreme outliers.

# 5b. At the county level: comparing all counties
```{r}
DataCounty = DataTotal[,.(
  PM2_5 = mean(PM2_5, na.rm = TRUE)
), by = c("COUNTY","Year")]

ggplot(DataCounty, aes(x=COUNTY, y=PM2_5)) +
  geom_point(aes(color= Year)) +
  theme(axis.text.x = element_text(angle = 90))


```
For most counties, we see that the average Yearly PM2.5 Concentration was higher in 2004 than 2019. Therefore, it can be said that the average yearly PM2.5 concentration decreased in most counties from 2004 to 2019. Counties where this was not the case were Del Norte, Mono, Siskiyou, and Trinity. This graph also shows that there was no 2004 data for the counties of Glenn, Madera, Napa, and Tehama.

# 5c. Within Los Angeles
```{r}
# isolated all LA County data so i could find a site
DataLA = DataTotal[COUNTY == "Los Angeles"]
unique(DataLA$city)
DataPasadena= DataLA[city == "Pasadena"]
unique(DataPasadena$`Site ID`)
# only one site in Pasadena, so I'll look at that one

ggplot(DataPasadena, aes(x=1, y=PM2_5)) +
  geom_jitter(aes(color= Year)) + 
  facet_grid(~Year) +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

mean(DataPasadena[Year == 2004]$PM2_5)
mean(DataPasadena[Year == 2019]$PM2_5)
```
The average PM2.5 concentration almost halved in Pasadena between from 2004 to 2019. This can be seen visually in the plotting of all observations in Pasadena during these two years. 
