---
title: "Assignment2"
author: "Sylvia Baeyens"
date: "10/7/2021"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages, echo= FALSE, include= FALSE}
#including necessary libraries
library(data.table)
library(dplyr)
library(tidyverse)
library(leaflet)
library(R.utils)
library(ggplot2)
```

## Loading Data Files & Merging
```{r cache= TRUE}
# Reading individual data set
individual <- data.table::fread("chs_individual.csv")

# Reading the regional data
regional <- data.table::fread("chs_regional.csv")

# Merging by location
MData <- merge(
  x = individual,
  y = regional,
  all.x = TRUE, all.y = FALSE
)
#the only common column was that of townname, don't need to specify a by.x/by.y
```

# Data Wrangling

## 1. Data Cleanup

After merging the data, make sure you don’t have any duplicates by counting the number of rows. Make sure it matches. In the case of missing values, impute data using the average within the variables “male” and “hispanic.” If you are interested (and feel adventurous) in the theme of Data Imputation, take a look at this paper on “Multiple Imputation” using the Amelia R package here.

```{r}
dim(individual)
dim(MData)
# both have 1200 rows, so there are no duplicates

summary(is.na(MData))
#there is missing data for agepft, height, weight, bmi, asthma, father_asmtha, mother_asthma, wheeze, hayfever, allergy, educ_parent, smoke, gasstove,fev, fvc, mmef, no_24hr, and pm 2_5_fr 

# imputing for bmi, smoke, gasstove & fev based on gender & race as these are the variables we will be inspecting later

MData[, bmi := fcoalesce(bmi, mean(bmi, na.rm = TRUE)),
    by = .(male, hispanic)]
MData[, fev := fcoalesce(fev, mean(fev, na.rm = TRUE)),
    by = .(male, hispanic)]

# could not impute for gasstove & smoke because they are binary
```


## 2. New Obesity Level variable

Create a new categorical variable named “obesity_level” using the BMI measurement (underweight BMI<14; normal BMI 14-22; overweight BMI 22-24; obese BMI>24). To make sure the variable is rightly coded, create a summary table that contains the minimum BMI, maximum BMI, and the total number of observations per category.

```{r}
MData[bmi< 14,obesity_level:="underweight"]
MData[bmi>= 14 & bmi< 22,obesity_level:="normal"]
MData[bmi>= 22 & bmi< 24,obesity_level:="overweight"]
MData[bmi>= 24,obesity_level:="obese"]

MData[, .(
    min_BMI = min(bmi, na.rm=TRUE),
    max_BMI = max(bmi, na.rm=TRUE),
    total_num = .N
    ),
    by = obesity_level
    ] %>% head(n = 4)
```


## 3. New exposure variable

Create another categorical variable named “smoke_gas_exposure” that summarizes “Second Hand Smoke” and “Gas Stove.” The variable should have four categories in total.

```{r}
MData[smoke == 0 & gasstove == 0,smoke_gas_exposure:="no exposure"]
MData[smoke == 1 & gasstove == 0,smoke_gas_exposure:="smoke exposure"]
MData[smoke == 0 & gasstove == 1,smoke_gas_exposure:="gas exposure"]
MData[smoke == 1 & gasstove == 1,smoke_gas_exposure:="smoke & gas exposure"]

MData[,.(
  obs= .N),
  by = smoke_gas_exposure
]%>% head(n = 5)
```


## 4. Summary Tables

Create four summary tables showing the average (or proportion, if binary) and sd of “Forced expiratory volume in 1 second (ml)” and asthma indicator by town, sex, obesity level, and “smoke_gas_exposure.”

```{r}

MData[, .(
    average_FEV = mean(fev, na.rm=TRUE),
    standarddev_FEV = sd(fev, na.rm=TRUE),
    average_asthmaindic = mean(asthma, na.rm=TRUE),
    standarddev_asthmaindic = sd(asthma, na.rm=TRUE)
    ),
    by = townname
    ] %>% knitr::kable(caption = "Average FEV & Asthma Indicator by Town")

MData[, .(
    average_FEV = mean(fev, na.rm=TRUE),
    standarddev_FEV = sd(fev, na.rm=TRUE),
    average_asthmaindic = mean(asthma, na.rm=TRUE),
    standarddev_asthmaindic = sd(asthma, na.rm=TRUE)
    ),
    by = male
    ] %>% knitr::kable(caption = "Average FEV & Asthma Indicator by Sex")

MData[, .(
    average_FEV = mean(fev, na.rm=TRUE),
    standarddev_FEV = sd(fev, na.rm=TRUE),
    average_asthmaindic = mean(asthma, na.rm=TRUE),
    standarddev_asthmaindic = sd(asthma, na.rm=TRUE)
    ),
    by = obesity_level
    ] %>% knitr::kable(caption = "Average FEV & Asthma Indicator by Obesity Level")

MData[, .(
    average_FEV = mean(fev, na.rm=TRUE),
    standarddev_FEV = sd(fev, na.rm=TRUE),
    average_asthmaindic = mean(asthma, na.rm=TRUE),
    standarddev_asthmaindic = sd(asthma, na.rm=TRUE)
    ),
    by = smoke_gas_exposure
    ] %>% knitr::kable(caption = "Average FEV & Asthma Indicator by Smoke & Gas Exposure")
```

# Looking at the Data- EDA 

The three questions of interest are:
  1. What is the association between BMI and FEV (forced expiratory volume)?
  2. What is the association between smoke and gas exposure and FEV?
  3. What is the association between PM2.5 exposure and FEV?

## EDA Checklist

Checking Dimensions & headers/footers
```{r}
dim(MData)
head(MData)
tail(MData)
```

Checking Variable Types & Taking Closer Look at Key Variables
```{r}
str(MData)

summary(MData$bmi)
summary(MData$fev)
summary(MData$pm25_mass)
```

Summary Statistics
```{r}
MData[, .(
    min_fev = min(fev, na.rm=TRUE),
    max_fev = max(fev, na.rm=TRUE),
    avg_fev = mean(fev, na.rm=TRUE),
    total_num = .N
    ),
    by = obesity_level
    ] %>% knitr::kable(caption = "FEV Summary Stats by Obesity Level")

MData[, .(
    min_fev = min(fev, na.rm=TRUE),
    max_fev = max(fev, na.rm=TRUE),
    avg_fev = mean(fev, na.rm=TRUE),
    total_num = .N
    ),
    by = smoke_gas_exposure
    ] %>% knitr::kable(caption = "FEV Summary Stats by Smoke & Gas Exposure")
```


# Data Visualization

## 1. Scatter plot comparing BMI vs FEV by townname
```{r}
ggplot(MData, aes(x=bmi, y=fev)) +
  geom_point() +
  facet_wrap(~townname) +
  geom_smooth(method ="lm") +
  ggtitle("BMI vs FEV in different CA towns")

```
It does appear from the above graphs that there is a positive relationship between increasing bmi and increasing FEV. As an individuals bmi increases, so, too, does their lung capacity. For all towns, the slope of the best fit line is greater than 0. This trend seems most apparent in the towns of Upland, Lake Elsinore, and Atascadero, as shown by the slopes of these best fit lines.

## 2. Stacked histograms of FEV by BMI category and FEV by smoke/gas exposure.
```{r}
ggplot(MData, aes(x=fev)) +
  geom_histogram(aes(fill=obesity_level))+
  scale_fill_brewer(palette = "Pastel1") +
  ggtitle("Visualizing FEV distribution by Obesity Levels")

ggplot(MData, aes(x=fev)) +
  geom_histogram(aes(fill=smoke_gas_exposure))+
  scale_fill_brewer(palette = "Pastel2") +
  ggtitle("Visualizing FEV distribution by Smoke and Gas Exposure Levels")
```
When looking at obesity level, the distribution of fev is normal for the normal weight group. However, for the obese & overweight groups, the data is skewed to the right whereas for the underweight group, the data is skewed to the left. When viewing this graph, we can assume that the average fev for the obese and overweight groups is greater than that of the normal group, and that the average fev for the underweight group is less than that of the normal group. 

We do not see a similar trend in the histogram showing fev by smoke & gas exposure level. For all of these groups, the fev distribution is normal, and one can assume that the average fev for all 4 groups is the same. 

## 3. Barchart of BMI by smoke/gas exposure.
```{r}
ggplot(MData) +
  geom_bar(mapping =aes(x=obesity_level, fill=smoke_gas_exposure)) +
  ggtitle("Visualizing Gas & Smoke Exposure Level counts by Obesity Levels")
```
Gas exposure is the most common form of exposure experienced by all 4 obesity level groups. The normal group seems to have the highest percentage of individuals experiencing no exposure. Besides that, the proportions of exposure levels appear to be the same among the different groups, showing that there might not be that large of a relationship between the type of exposure experienced and obesity level.

## 4. Statistical summary graphs of FEV by BMI and FEV by smoke/gas exposure category
```{r}
ggplot(MData, aes(x= obesity_level, y= fev)) +
  stat_summary(fun.min = min, fun.max= max, fun = median) +
  ggtitle("Visualizing Range & Median of FEV by Obesity Levels")

ggplot(MData, aes(x= smoke_gas_exposure, y= fev)) +
  stat_summary(fun.min = min, fun.max= max, fun = median) +
  ggtitle("Visualizing Range & Median of FEV by Obesity Levels")
```
The average fev is greatest for the obese and overweight groups at about 2250, and lowest for the underweight group at 1700. The normal group has the largest range of fev values and an average fev of 2000. The normal and obese groups have the greatest maximum fev values, while the normal and underweight groups have the smallest mimimum values. It can be seen that there is some correlation between fev and obesity level, and that there is a greater likelihood to have a large fev value at a high bmi. This could be because a high bmi leads to a larger frame and therefore higher lung capacity. An underweight individual likely has less room within their body for their lungs to expand, and therefore, has less lung capacity compariatively. This trend was shown in the earlier graph visualizing bmi and fev by town.


## 5. A leaflet map showing the concentrations of PM2.5 mass in each of the CHS communities.
```{r eval = knitr::is_html_output(excludes = "gfm")}

temp.pal = colorNumeric(c('darkgreen','yellow','red'), domain = MData$pm25_mass)

leaflet(MData) %>%
  addProviderTiles('CartoDB.Positron') %>%
  addCircles(
    lat = ~lat, lng=~lon, 
    color = ~ temp.pal(pm25_mass),
    opacity=1, fillOpacity=1
    )  %>%
  addLegend('bottomleft', pal=temp.pal, values=MData$pm25_mass,
          title='PM2.5 Mass', opacity=1)
#the leaflet graph won't print in knit github document
```
The mass of PM2.5 Particles is greatest in the town of Mira Loma, east of LA. The other towns near the Riverside/San Bernardino areas also have large PM 2.5 Mass. As you move further away from the LA metropolitan area, the mass of the averag PM2.5 particles decreases. The smallest average particle mass is found in San Luis Obispo. This is likely due to the urban and industrial nature of the LA metropolitan area and its surroundings, whereas SLO has less manufacturing and heavy industrial plants, leading to less large air pollutants.

## 6. Choose a visualization to examine whether PM2.5 mass is associated with FEV.
```{r}
ggplot(MData, aes(x=pm25_mass, y=fev)) +
  geom_point(aes(color=  townname)) +
  geom_smooth(method= "lm") +
  ggtitle("Visualizing the Association between FEV & PM2.5 Particle Mass")
```
This final graph visualizes fev by pm 2.5 particle mass. Each town receives a unique color, which goes in hand with the above leaflet graph, confirming that Mira Loma, Upland, and Riverside observed the largest PM 2.5 Particles, while Lompoc and Santa Maria had the smallest Pm 2.5 Particles. There is a slight negative relationshop between the two variables. As the mass of the PM 2.5 particles increases, the average fev decreases. This could imply that breathing in the larger air pollutant particles could have a negative effect on lung function; however, this effect is very small and almost negligible. 
