---
title: "Assignment 4- HPC & SQL"
author: "Sylvia Baeyens"
date: "due 11/19/2021"
output:
  github_document: 
    html_preview: false
  html_document: default
  word_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages, echo= FALSE, include= FALSE}
#including necessary libraries
library(tidyverse)
library(tidytext)
library(dplyr)
library(ggplot2)
library(data.table)
library(httr)
```

# HPC
## Problem 1: Make sure your code is nice

Rewrite the following R functions to make them faster. It is OK (and recommended) to take a look at Stackoverflow and Google
```{r rewriting functions}

# Total row sums
fun1 <- function(mat) {
  n <- nrow(mat)
  ans <- double(n) 
  for (i in 1:n) {
    ans[i] <- sum(mat[i, ])
  }
  ans
}

fun1alt <- function(mat) {
  ans = rowSums(mat)
  ans
}

# Cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

fun2alt <- function(mat) {
 ans= t(apply(mat, 1, cumsum))
 ans
}


# Use the data with this code
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)

# Test for the first
microbenchmark::microbenchmark(
  fun1(dat),
  fun1alt(dat), unit = "relative", check = "equivalent"
)

# Test for the second
microbenchmark::microbenchmark(
  fun2(dat),
  fun2alt(dat), unit = "relative", check = "equivalent"
)

```
In both cases, our alternative code, which contaisn no for loops, runs much faster.


## Problem 2: Make things run faster with parallel computing

The following function allows simulating PI
```{r 2a}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132
```

In order to get accurate estimates, we can run this function multiple times, with the following code:

```{r 2b}
# This runs the simulation a 4,000 times, each with 10,000 points
set.seed(1231)
system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans))
})
```

Rewrite the previous code using parLapply() to make it run faster. Make sure you set the seed using clusterSetRNGStream():

```{r 2c}

library(parallel)

system.time({
  cl <- makePSOCKcluster(1L)
  clusterSetRNGStream(cl, 1231)
  clusterExport(cl, c("sim_pi"), envir = environment())
  ans = unlist(parLapply(cl = cl, 1:4000, sim_pi, n = 10000))
  print(mean(ans))
  stopCluster(cl)
  ans
})
```
Parallizing the code makes it run much faster.


# SQL

Setup a temporary database by running the following chunk
```{r}
# install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
film <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film.csv")
film_category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film_category.csv")
category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/category.csv")

# Copy data.frames to database
dbWriteTable(con, "film", film)
dbWriteTable(con, "film_category", film_category)
dbWriteTable(con, "category", category)
```

## Question 1
How many many movies is there avaliable in each rating catagory.
```{sql connection=con}

SELECT rating,
  COUNT (*) as Number_Movies
FROM film
GROUP BY rating

```


## Question 2
What is the average replacement cost and rental rate for each rating category.
```{sql connection=con}

SELECT rating,
  AVG(replacement_cost) as AvgReplacementCost, 
  AVG(rental_rate) as AvgRentalRate
FROM film
GROUP BY rating

```

## Question 3
Use table film_category together with film to find the how many films there are with each category ID
```{sql connection=con}

SELECT category_id,
  COUNT (*) as Number_Films
FROM film_category
GROUP BY category_id

```

## Question 4 

Incorporate table category into the answer to the previous question to find the name of the most popular category.
```{sql connection=con}

SELECT *
  FROM category as a
  INNER JOIN 
    (SELECT category_id, COUNT (*) as Number_Films
      FROM film_category GROUP BY category_id) as b
  ON a.category_id = b.category_id
  ORDER BY Number_Films DESC
  LIMIT 1

```
The most popular movie category is the sports category. 

## Clean up
```{r}
# clean up
dbDisconnect(con)
```